---
title: "Differential discovery with `CATALYST`"
date: "`r BiocStyle::doc_date()`"
author:
- name: Helena L Crowell
  affiliation:
  - &IMLS Institute for Molecular Life Sciences, University of Zurich, Switzerland
  - &SIB SIB Swiss Institute of Bioinformatics, University of Zurich, Switzerland
  email: helena.crowell@uzh.ch
- name: Mark D Robinson
  affiliation:
  - *IMLS
  - *SIB
package: "`r BiocStyle::pkg_ver('CATALYST')`"
bibliography: "`r file.path(system.file('extdata', package = 'CATALYST'), 'refs.bib')`"
vignette: >
  %\VignetteIndexEntry{"2. Differential discovery"}
  %\VignettePackage{CATALYST}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

---

<span style="color:firebrick">**Most of the pipeline and visualizations presented herein have been adapted from @Nowicka2019-F1000's *"CyTOF workflow: differential discovery in high-throughput high-dimensional cytometry datasets"* available [here](https://f1000research.com/articles/6-748/v4).**</span>

```{r warning = FALSE, message = FALSE}
# load required packages
library(CATALYST)
library(cowplot)
library(flowCore)
library(diffcyt)
library(scater)
library(SingleCellExperiment)
```

# Example data

- `PBMC_fs`:  
a `flowSet` holding PBMCs samples from 4 patients, each containing between 500 and 1000 cells. For each sample, the expression of 10 cell surface and 14 signaling markers was measured before (REF) and upon BCR/FcR-XL stimulation (BCRXL) with B cell receptor/Fc receptor crosslinking for 30', resulting in a total of 8 samples.  
This data set represents a subset of data originating from @Bodenmiller2012 that was also used in the [citrus](https://github.com/nolanlab/citrus) paper [@Bruggner2014-CITRUS].
- `PBMC_panel`:  
a data.frame containing each marker's column name in the FCS file (`fcs_colname` column), its targeted protein marker (`antigen` column), and the `marker_class` ("type" or "state"). 
- `PBMC_md`:  
a data.frame where rows correspond to samples, and columns specify each sample's `file_name`, `sample_id`, `condition`, and `patient_id`.

```{r load-data}
# load example data
data(PBMC_fs, PBMC_panel, PBMC_md)
PBMC_fs
head(PBMC_panel)
head(PBMC_md)
```

The code snippet below demonstrates how to construct a `flowSet` from a set of FCS files. However, we also give the option to directly specify the path to a set of FCS files (see next section).

```{r eval=FALSE}
# download exemplary set of FCS files
url <- "http://imlspenticton.uzh.ch/robinson_lab/cytofWorkflow"
zip <- "PBMC8_fcs_files.zip"
download.file(paste0(url, "/", zip), destfile = zip, mode = "wb")
unzip(zip)

# read in FCS files as flowSet
fcs <- list.files(pattern = ".fcs$")
fs <- read.flowSet(fcs, transformation = FALSE, truncate_max_range = FALSE)
```

# Data preparation

Data used and returned throughout differential analysis are held in objects of the `r BiocStyle::Biocpkg("SingleCellExperiment")` class. To bring the data into the appropriate format, `prepData()` requires the following inputs:

- `x`: a `flowSet` holding the raw measurement data, or a character string that specifies a path to a set of FCS files.
- `panel`: a 2 column data.frame that contains for each marker of interest i) its column name in the raw input data, and ii) its targeted protein marker.
- `md`: a data.frame with columns describing the experimental design.

Optionally, `features` will specify which columns (channels) to keep from the input data. Here, we keep all measurement parameters (default value `features = NULL`). 

```{r prepData}
(sce <- prepData(PBMC_fs, PBMC_panel, PBMC_md))
```

We provide flexibility in the way the panel and metadata table can be set up. Specifically, column names are allowed to differ from the example above, and multiple factors (patient ID, conditions, batch etc.) can be specified. Arguments `panel_cols` and `md_cols` should then be used to specify which columns hold the required information. An example is given below:

```{r eval=FALSE}
# alter panel column names
panel2 <- PBMC_panel
colnames(panel2)[1:2] <- c("channel_name", "marker")

# alter metadata column names & add 2nd condition
md2 <- PBMC_md
colnames(md2) <- c("file", "sampleID", "cond1", "patientID")
md2$cond2 <- rep(c("A", "B"), 4)

# construct SCE
prepData(PBMC_fs, panel2, md2, 
    panel_cols = list(channel = "channel_name", antigen = "marker"),
    md_cols = list(file = "file", id = "sampleID", 
        factors = c("cond1", "cond2", "patientID")))
```

Note that, independent of the input panel and metadata tables, the constructor will fix the names of mandatory slots for latter data accession (`sample_id` in the `rowData`, `channel_name` and `marker_name` in the `colData`). The `md` table will be stored under `experiment_info` inside the `metadata`.

# Diagnostic plots

## `plotCounts`: Number of cells measured per sample

The number of cells measured per sample may be plotted with `plotCounts`. This plot should be used as a guide together with other readouts to identify samples where not enough cells were assayed. Here, the grouping of samples (x-axis) is controlled by `group_by`; bars can be colored by a an additional cell metadata variable (argument `color_by`):

```{r plotCounts-1, fig.width = 5, fig.height = 3}
plotCounts(sce, 
    group_by = "sample_id", 
    color_by = "condition")
```

As opposed to plotting absolute cell counts, argument `prop` can be used to visualize relative abundances (frequencies) instead:

```{r plotCounts-2, fig.width = 4, fig.height = 3}
plotCounts(sce, 
    prop = TRUE,
    group_by = "condition", 
    color_by = "patient_id")
```

## `pbMDS`: Pseudobulk-level multi-dimensional scaling plot

A multi-dimensional scaling (MDS) plot on median expresion values may be rendered with `plotMDS`. Such a plot will give a sense of similarities between samples in an unsupervised way and of key difference in expression before conducting any formal testing. In our example, we can see a clear separation between reference (REF) and stimulation condition (BCRXL).

```{r pbMDS, fig.width = 5}
pbMDS(sce, 
    color_by = "condition", 
    label_by = "sample_id",
    shape_by = "patient_id")
```

## `plotExprHeatmap`: Heatmap of (scaled) median marker expressions

`plotExprHeatmap` will show a heatmap on median marker intensities with hierarchically clustered rows (samples) and columns (markers). This plot should give an idea of which markers will drive sample clustering, and how similiar samples are in their expression profile. We specify `bin_anno = TRUE` to display expression values inside each bin, and `row_anno = TRUE` to include row annotations for each factor in `metadata(daf)`. 

```{r plotExprHeatmap, fig.width = 10, fig.height = 4}
plotExprHeatmap(sce, row_anno = TRUE, bin_anno = FALSE)
```

# Clustering

## `cluster`: *FlowSOM* clustering & *ConsensusClusterPlus* metaclustering

`r BiocStyle::Biocpkg("CATALYST")` provides a simple wrapper to perform high resolution `FlowSOM` clustering and lower resolution `ConsensusClusterPlus` metaclustering. By default, the data will be initially clustered into `xdim = 10` x `ydim = 10` = 100 groups. Secondly, the function will metacluster populations into 2 through `maxK` (default 20) clusters. To make analyses reproducible, the random seed may be set via `seed`. By default, if the `colData(sce)$marker_class` column is specified, the set of markers with marker class `"type"` will be used for clustering (argument `features = "type"`). Alternatively, the markers that should be used for clustering can be specified manually.

```{r cluster}
sce <- cluster(sce, features = "type", 
    xdim = 10, ydim = 10, maxK = 20, 
    verbose = FALSE, seed = 1)       
```

Let K = `xdim` x `ydim` be the number of `r BiocStyle::Biocpkg("FlowSOM")` clusters. `cluster` will add information to the following slots of the input `SingleCellExperiment`: 

- `rowData`:
  - `cluster_id`: cluster ID as inferred by `r BiocStyle::Biocpkg("FlowSOM")`. One of 1, ..., K.
- `colData`:
  - `marker_class`: factor `"type"` or `"state"`. Specifyies whether a marker has been used for clustering or not, respectively.
- `metadata`:
  - `SOM_codes`: a table with dimensions K x (# type markers). Contains the SOM codes. 
  - `cluster_codes`: a table with dimensions K x (`maxK` + 1). Contains the cluster codes for all metaclusterings.
  - `delta_area`: a `ggplot` object (see below for details).

## Delta area plot

The delta area represents the amount of extra cluster stability gained when clustering into k groups as compared to k-1 groups. It can be expected that high stability of clusters can be reached when clustering into the number of groups that best fits the data. The "natural" number of clusters present in the data should thus corresponds to the value of k where there is no longer a considerable increase in stability (pleateau onset). For more details, the user can refer to the original description of the consensus clustering method [@Monti2003-ConsensusClusterPlus].

```{r delta-area, fig.width = 5, fig.height = 2.5}
# access & render delta area plot
# (equivalent to metadata(sce)$delta_area)
delta_area(sce)
```

## `plotMedExprs`: Median marker-expressions by cluster

A combined boxplot and jitter of median marker intensitied can be generated via `plotMedExprs`. Here, argument `features` (default `"state"`, which is equvalent to `state_markers(sce)`) controls which markers to include. `features = NULL` will include all markers (and is equivalent to `rownames(sce)`).

In order to compare medians for each cluster, and potentially identify changes across conditions early on, we specify `facet = "cluster_id"`: 

```{r plotMedExprs-1, fig.width = 12, fig.height = 5}
p <- plotMedExprs(sce, facet = "cluster_id", k = "meta8")
p$facet$params$ncol <- 4; p
```

Alternatively, we can facet the above plot by `antigen` in order to compare marker expressions calculated over all cells across conditions:

```{r plotMedExprs-2, fig.width = 12, fig.height = 4}
p <- plotMedExprs(sce, facet = "antigen")
p$facet$params$ncol <- 7; p
```

## `plotClusterExprs`: Marker-densities by cluster

Distributions of marker intensities (arcsinh-transformed) across cell populations of interest can be plotted with `plotClusterExprs`. We specify `features = "type"` (equivalent to `type_markers(sce)`), to include type-markers only. Here, blue densities (top row) are calculated over all cells and serve as a reference.

```{r plotClusterExprs, message = FALSE, fig.width = 12, fig.height = 8}
plotClusterExprs(sce, k = "meta8", features = "type")
```

## `mergeClusters`: Manual cluster merging

Provided with a 2 column data.frame containing `old_cluster` and `new_cluster` IDs, `mergeClusters` allows for manual cluster merging of any clustering available within the input `SingleCellExperiment` (i.e. the `xdim` x `ydim` `r BiocStyle::Biocpkg("FlowSOM")` clusters, and any of the 2-`maxK` `r BiocStyle::Biocpkg("ConsensusClusterPlus")` metaclusters). For latter accession (visualization, differential testing), the function will assign a unique ID (specified with `id`) to each merging, and add a column to the `cluster_codes` inside the `metadata` slot of the input `SingleCellExperiment`. 

```{r mergeClusters}
data(merging_table)
head(merging_table)
sce <- mergeClusters(sce, k = "meta20", table = merging_table, id = "merging1")
head(cluster_codes(sce))[, seq_len(10)]
```

## `plotClusterHeatmap`: Heatmap of (meta)clustering results

Clusterings and metaclusters maybe be viewing with the `plotClusterHeatmap`. In its 1st panel, the function will display median (arcsinh-transformed and *optionally* scaled) cell-type marker expressions (across all samples). Depending on argument `hm2`, the 2nd panel will vary as follows:

- `"abundances"`: cluster frequencies by sample;
- `"state"`: median cell state marker expressions across clusters  
(analogous to the left-hand side heatmap);
- a character string/vector corresponding to one/multiple marker(s):  
median marker expressions by sample.

Argument `scale` (default `TRUE`) specifies whether scaled values should be plotted. These correspond to arcsinh-transformed expression values scaled between 0 and 1 using low (1%) and high (99%) percentiles as boundaries. Note that, in any case, hierarchical clustering is performed on the unscaled data.  
While it is only used here for visualization, this additional transformation of the arcsinh-transformed data can sometimes give a better representation of relative differences in marker expression between cell populations.

```{r plotClusterHeatmap-1, fig.width = 8, fig.height = 6}
# median pS6 expression by sample as 2nd heatmap
plotClusterHeatmap(sce, hm2 = "pS6", k = "meta12", m = "meta6")
```

```{r plotClusterHeatmap-2, fig.width = 10, fig.height = 6}
# population frequencies by sample as 2nd heatmap
plotClusterHeatmap(sce, hm2 = "abundances", 
    draw_freqs = TRUE, row_anno = FALSE)
```

## `plotAbundances`: Relative population abundances

Relative population abundances for any clustering of interest can be plotted with `plotAbundances`. Argument `by` will specify whether to plot proportions for each sample or cluster; `group_by` determines the grouping within each panel as well ascolor coding.

- If `by = "sample_id"`, the function displays each sample's cell type composition, and the size of a given stripe reflects the proportion of the corresponding cell type the given sample. Argument `group_by` then specifies the facetting.
- If `by = "cluster_id"`, argument `group_by` then specifies the grouping and color coding. 

```{r plotAbundances, fig.width = 6, fig.height = 4}
plotAbundances(sce, k = "meta12", by = "sample_id", group_by = "condition")
plotAbundances(sce, k = "merging1", by = "cluster_id", 
    group_by = "condition", shape_by = "patient_id")
```

# Dimensionality reduction

The number of cells in cytometry data is typically large, and for visualization of cells in a two-dimensional space it is often sufficient to run dimension reductions on a subset of the data. Thus, `CATALYST` provides the wrapper function `runDR` to apply any of the dimension reductions available from `BiocStyle::Biocpkg("scater")` using

1. the subset of features specified via argument `features`; either a subset of `rownames(.)` or, e.g., `"type"` for `type_markers(.)` (if `marker_classes(.)` have been specified).
2. the subset of cells specified via argument `cells`; either `NULL` for all cells, or `n` to sample a random subset of n cells per sample.

To make results reproducible, the random seed should be set via `set.seed` *prior* to computing reduced dimensions:

```{r runDR}
set.seed(1601)
sce <- runDR(sce, dr = "UMAP", cells = 500, features = "type")
```

Alternatively, dimension reductions can be computed using one of `r BiocStyle::Biocpkg("scater")`'s `runX` functions (`X = "TSNE", "UMAP", ...`). Note that, by default, `scater` expects expression values to be stored in the `logcounts` assay of the SCE; specification of `exprs_values = "exprs"` is thus required:

```{r runUMAP-scater, eval = FALSE}
sce <- runUMAP(sce, exprs_values = "exprs")
```

DRs available within the SCE can be viewed via `reducedDimNames` and accessed with `reducedDim(s)`:

```{r}
# view & access DRs
reducedDimNames(sce)
head(reducedDim(sce, "UMAP"))
```

While `r BiocStyle::Biocpkg("scater")`'s `plotReducedDim` function can be used to visualize DRs, `CATALYST` provides the `plotDR` wrapper, specifically to allow for coloring cells by the various clusterings available, and to support facetting by metadata factors (e.g., experimental condition, sample IDs):

```{r plotDR-1, fig.width = 8, height = 6}
# color by marker expression & split by condition
plotDR(sce, color_by = c("pS6", "pNFkB"), facet_by = "condition")
```

```{r plotDR-2, fig.width = 8, fig.height = 5}
# color by 8 metaclusters & split by sample ID
p <- plotDR(sce, color_by = "meta8", facet_by = "sample_id")
p$facet$params$ncol <- 4; p
```

# Filtering

SCEs constructed with `prepData` can be filtered using the `filterSCE` function, which allows for filtering of both cells and markers according to conditional statements in `dplyr`-style. When filtering on `cluster_id`s, argument `k` specifies which clustering to use (the default `NULL` uses `colData` column `"cluster_id"`). Two examples are given below:

```{r filterSCE, fig.width = 8, fig.height = 3}
u <- filterSCE(sce, patient_id == "Patient1")
table(u$sample_id)

u <- filterSCE(sce, k = "meta8",
    cluster_id %in% c(1, 3, 8))
plot_grid(
    plotDR(sce, color_by = "meta8"),
    plotDR(u, color_by = "meta8"))
```

# Differental testing with `r BiocStyle::Biocpkg("diffcyt")`

`r BiocStyle::Biocpkg("CATALYST")` has been designed to be compatible with the `r BiocStyle::Biocpkg("diffcyt")` package [@Weber2019-diffcyt], which implements statistical methods for differential discovery in high-dimensional cytometry (including flow cytometry, mass cytometry or CyTOF, and oligonucleotide-tagged cytometry) using high-resolution clustering and moderated tests. The input to the `r BiocStyle::Biocpkg("diffcyt")` pipeline can either be raw data, or a `SingleCellExperiment` object. We give an exmaple of the latter below.  
Please refer to the `r BiocStyle::Biocpkg("diffcyt")` [vignette](https://bioconductor.org/packages/3.7/bioc/vignettes/diffcyt/inst/doc/diffcyt_workflow.html) and R documentation (`??diffcyt`) for more detailed information.

```{r diffcyt, message = FALSE, warning = FALSE, fig.show = "hide"}
# create design & constrast matrix
design <- createDesignMatrix(ei(sce), cols_design = "condition")
contrast <- createContrast(c(0, 1))

# test for
# - differential abundance (DA) of clusters
# - differential states (DS) within clusters
res_DA <- diffcyt(sce, clustering_to_use = "meta10",
    analysis_type = "DA", method_DA = "diffcyt-DA-edgeR",
    design = design, contrast = contrast, verbose = FALSE)
res_DS <- diffcyt(sce, clustering_to_use = "meta10",
    analysis_type = "DS", method_DS = "diffcyt-DS-limma",
    design = design, contrast = contrast, verbose = FALSE)
```

## `plotDiffHeatmap`: Heatmap of differential testing results

Differential testing results returned by `r BiocStyle::Biocpkg("diffcyt")` can be displayed with the `plotDiffHeatmap` function.

For differential abundance (DA) tests, `plotDiffHeatmap` will display

- median (arcsinh-transformed) cell-type marker expressions (across all samples), and
- relative cluster abundances by samples

For differential state (DS) tests, `plotDiffHeatmap` will display

- median (arcsinh-transformed) cell-type marker expressions (across all samples)
- median (arcsinh-transformed) cell-state marker expressions by sample

Clusters (DA) and cluster-marker combinations (DS), respectively, will be marked as *significant* if their adjusted p-value falls below the threshold value specified with `th` (default 0.1), and will be ordered by significance if `order = TRUE` (the default). The number of top findings to display can be specified with `top_n` (default 20). When `normalize = TRUE`, the right-hand side heatmap will display Z-score normalized values. For DA, cluster frequencies will be arcsine-square-root scaled prior to normalization.

```{r plotDiffHeatmap-da, fig.width = 8, fig.height = 4}
plotDiffHeatmap(sce, res_DA, all = TRUE, th = 0.05)
```

```{r plotDiffHeatmap-ds, fig.width = 7, fig.height = 5}
plotDiffHeatmap(sce, res_DS, hm1 = FALSE, top_n = 15)
```

# More

## Using other clustering algorithms

While `r BiocStyle::Biocpkg("FlowSOM")` has proven to perform well in systematic comparisons of clustering algorithms for CyTOF data [@Weber2016-clustering; @Freytag2018-clustering], it is not the only method out there. Here we demonstrate how clustering assignments from another clustering method, say, `r Githubpkg("JinmiaoChenLab", "Rphenograph")`, could be incorporated into the SCE to make use of the visualizations available in `CATALYST`. Analogous to the example below, virtually any clustering algorithm could be applied, however, with the following limitation:  
The `ConsensusClusterPlus` metaclusterings applied to the initial `FlowSOM` clustering by `CATALYST`'s `cluster` function have a hierarchical cluster structure. Thus, clustering IDs can be matched from a higher resolution (e.g. 100 SOM clusters) to any lower resolution (e.g., 2 through 20 metaclusters). This is not guaranteed for other clustering algorithms. Thus, we store only a single resolution in the cell metadata column `cluster_id`, and a single column under `metadata` slot `cluster_codes` containing the unique cluster IDs. Adding additional resolutions to the `cluster_codes` will fail if cluster IDs can not be matched uniquely across conditions, which will be the case for any non-hierarchical clustering method.

```{r other-clusterings, message = FALSE, warning = FALSE, fig.width = 5, fig.height = 2.5}
# subset type-marker expression matrix
es <- assay(sce, "exprs")
es <- es[type_markers(sce), ]

# run clustering method X
# (here, we just split the cells into 
# equal chunks according to CD33 expression)
cs <- split(seq_len(ncol(sce)), cut(es["CD33", ], nk <- 10))
kids <- lapply(seq_len(nk), function(i) {
    rep(i, length(cs[[i]]))
})
kids <- factor(unlist(kids))

# store cluster IDs in cell metadata & codes in metadata
foo <- sce
foo$cluster_id[unlist(cs)] <- unlist(kids)
metadata(foo)$cluster_codes <- data.frame(
    custom = factor(levels(kids), levels = levels(kids)))

# tabulate cluster assignments
table(cluster_ids(foo, "custom"))
# plot relative cluster abundances by sample
plotClusterHeatmap(foo, k = "custom")
```

## Customizing visualizations

Most of `CATALYST`'s plotting functions return `ggplot` objects whose aesthetics can (in general) be modified easily. However, while e.g. theme aesthetics and color scales can simply be added to the plot, certain modifications can be achieved only through overwriting elements stored in the object, and thus require a decent understanding of its structure.

Other functions (`plotExprHeatmap`, `plotClusterHeatmap` and `plotDiffHeatmap`) generate objects of the `Heatmap` of `HeatmapList` class from the `r BiocStyle::Biocpkg("ComplexHeatmap")` package, and are harder to modify once created. Therefore, `CATALYST` tries to expose a reasonable amount of arguments to the user that control key aesthetics such as the palettes used for coloring clusters and heatmaps.

The examples below serve to illustrate how some less exposed `ggplot` aesthetics can be modified in retrospect, and the effects of different arguments that control visualization of `r BiocStyle::Biocpkg("ComplexHeatmap")` outputs.

### Modifying `ggplot`s

```{r plotMedExprs-custom, fig.width = 8, fig.height = 2}
p <- plotMedExprs(sce, k = "meta4", facet_by = "cluster_id")
# facetting layout is 2x2; plot all side-by-side instead
p$facet$params$nrow <- 1
# remove points
p$layers <- p$layers[-1]
# overwrite default colors
p <- p + scale_color_manual(values = c("royalblue", "orange"))
# remove x-axis title, change angle & decrease size of labels
(p + labs(x = NULL) + theme(axis.text.x = element_text(angle = 90, size = 8)))
```

### Modifying `ComplexHeatmap`s

```{r plotClusterHeatmap-custom, fig.width = 6, fig.height = 3}
plotClusterHeatmap(sce,
    k = "meta8", 
    m = "meta4",
    hm2 = "abundances",
    # include dendrograms
    row_dend = TRUE, 
    col_dend = TRUE, 
    # primary & merging cluster palettes
    k_pal = hcl.colors(8, "Vik"),     
    m_pal = hcl.colors(4, "Tropic"), 
    # 1st & 2nd heatmap coloring
    hm1_pal = c("grey95", "blue"),  
    hm2_pal = c("grey95", "red3"))
```

```{r plotExprHeatmap-minimal, fig.width = 8, fig.height = 3}
# minimal heatmap
plotExprHeatmap(sce,
    row_anno = FALSE,   # don't annotate samples
    row_clust = FALSE,  # keep samples in original order
    col_clust = FALSE,  # keep markers in original order
    draw_freqs = FALSE, # don't include sample sizes
    bin_anno = FALSE,   # don't annotate bins
    hm_pal = c("grey95", "navy"))
```

```{r plotExprHeatmap-complete, fig.width = 12, fig.height = 4}
# complete heatmap
plotExprHeatmap(sce, row_anno = TRUE,   # annotate samples
    row_clust = TRUE, col_clust = TRUE, # cluster samples/markers
    row_dend = TRUE, col_dend = TRUE,   # include dendrograms
    draw_freqs = TRUE, # include barplot of sample sizes
    bin_anno = TRUE,   # annotate bins with value
    hm_pal = c("grey95", "orange"))
```

# Session information

```{r session-info}
sessionInfo()
```

# References